\documentclass[final,a4paper,12pt]{article}

\usepackage{notes}
\setlength\parskip{1ex}
\setlength\parindent{0cm}
\usepackage{float}
\usepackage{mathtools}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\newtheorem{problem}{Problem}[section]

\title{Geophysics III}
\author{Calculus Tutorial}
\date{}

\begin{document}
\maketitle
\tableofcontents

\section{Introduction}
Derivatives and integrals seem to frighten most students.  Perhaps because most
of us are taught to fear math and the further one gets in math the more it
should be feared.  Calculus can at times be rather complex, but in general it is
relatively easy (especially derivatives) and requires learning a few simple
rules and a handful of standard forms.  Most derivatives and many integrals can
be simplified into one of any of these standard forms, but it requires seeing
patterns that allow us to transform a more complicated forms into a simple
standard one.  Mostly, learning how to use calculus will take a little
experience but not a lot of hard work.

Calculus is the study of infinitesimals, that is the study of operations
performed on functions as changes in the distance between two knowns become
really really small. This allow us to determine slopes and inflection points,
areas and volumes, maxima and minima, and averages.  To understand calculus, we
first need to understand the concept of limits and infinitesimals.  We'll start
there, but then continue onto derivatives and integrals which give us all the
wonderful things I just mentioned.

\section{Limits}
Limits give us the ability to describe the class of problems where we wish to
know the value of a function while approaching some known point, but cannot
quite reach it.  This may sound a little silly at first blush, but there are
plenty of points we cannot reach, $\infty$ and $-\infty$ for starters.  But for
some functions, such as $x^{-1}$, we cannot evaluate the function at 0 or
another some other well defined point(s).  Don't believe me.  Go ahead.  Your
calculator will cowardly refuse to answer if you try.

We write a the limit of a function at $x$ as
\begin{equation}
    \mlbox[
        labelstyle={xshift=-3cm,yshift=0.5em},
        linestyle={out=270,in=90, latex-}]
        {\lim_{x \to x_0}} {indicates limit as $x$ approaches $x_0$}
    \mlbox[
        labelstyle={below of=m\themathLableNode,below of=m\themathLableNode}]
        {f(x)} {function of $x$}
    = 
    \mlbox[
        labelstyle={xshift=2cm},
        linestyle={out=270,in=90, latex-}]
        {L} {limit of function at $x_0$*} ,
\end{equation}
*where $f(x_0)$ may be undefined, but can be predicted.

So how do limits work?  As an example, suppose we wanted to know what
happens to $\exp(-x)$ as $x$ approaches 1.  This is a trivial case since we can
easily evaluate the function to find $\exp(-1) = e^{-1}$.

Now suppose we wanted to know what happens when $x$ becomes infinite.  We
cannot directly evaluate $\exp(-\infty)$ so we need a way to determine the value
at $\infty$, we write this problem as
\begin{equation}
    \lim_{x \to \infty} \exp(-x) = ? ,
\end{equation}
which we would read as, the limit of $\exp(-x)$ as $x$ approaches $\infty$.  If
we rewrite this problem just a little,
\begin{equation}
    \lim_{x \to \infty} \frac{1}{\exp(x)} = \frac{1}{\displaystyle \lim_{x \to \infty} \exp(x)} .
\end{equation}
We can write $\exp(x)$ as an infinite series
\begin{equation}
    \exp(x) = \sum_{k = 0}^\infty \frac{x^k}{k!} \approx 1 + x + \frac{x^2}{2} +
    \frac{x^3}{6} + \frac{x^4}{24} + \cdots .
\end{equation}
Clearly as $x \to \infty$, this infinite series then implies $\displaystyle \lim_{x \to \infty}
\exp(x) = \infty$ and likewise $\displaystyle \lim_{x \to \infty} \exp(-x) = 0$.  In
fact, one could prove that for any positive number greater than 1,
$a^{-\infty} = 0$.  The result detailed above really reiterates what we could have
guessed all along since the answer is well known, but it may not
always be possible to guess accurately---we'll generally stick to things that
are well known in this class.

Let's return to $x^{-1}$ as $x \to 0$.  If we were to plot this function, we
would see that for positive values of $x$, the function as small values is very
large and for small negative values it is very small.  This presents a problem
because the two branches of the function approach different values depending on
which side of 0 we approach from.  Hence, for some cases, we have to indicate
our direction of approach which we can do by $x \to 0^+$ from above and $x \to
0^-$ from below.

Like I mentioned before, we'll stick to cases where we know what the asymptotic
value is (or easy to guess) without having to do a bunch of rigorous work.
Hopefully now, you have an idea of what a limit is doing and how we write it.

\section{Infinitesimals}
Zeno, an ancient Greek mathematician, posed a paradox, that illustrates
infinitesimals fantastically.  In the course of a race, a runner will reach a
point that is half the distance between the start and finish.  Some time later,
he is half way between the half way mark and the finish (at the quarter mark).
And again he must reach a point that is half way between the quarter mark and
the finish---and so on, such that he must always reach a point that is half way
between a previous half way mark and the finish line without crossing.  So then,
how can the runner ever reach the finish line if they must always reach a point
half way between the finish and a point not quite there.  An infinitesimal is
just that, a point that you can never quite reach.  How small are they?  I don't
know, but I can always make a quantity smaller than a number you suggest.  It is
kinda like when kids discover infinity and find that some infinities are larger
than others.

``Oh yeah, infinity plus one.''

``Nuh-uh, infinity times infinity.''

Calculus relies on this concept of infinitesimals, small numbers so small such
that there is nothing smaller except 0 itself.


\section{Derivatives}
When we take a derivative we are determining how fast a function is changing
with respect to something---a spatial coordinate (say $z$), time or
whatever.  You are already familiar with derivatives.  Your car's speedometer
reports a derivative of your position with respect to time.  The slopes on a ski
run are the derivative of elevation with respect to position.  The cricket
scoring runs per over is also a derivative with respect to overs.  So
derivatives are used frequently even though we may not always think of them in
this way.

\subsection{Slope}
Derivatives are all about rates and differences.  More specifically, the rate in
change of a function at a point.  As an example, we will look at one of the
simplest functions one can draw, a line.

Recall the general equation of a line, $y = m x + b$, where $m$ is the slope and
$b$ is the $y$-intercept.  A derivative is very similar to our definition of
slope, which for a line is defined as 
\begin{equation}
    m = \frac{y_2 - y_1}{x_2 - x_1} ,
\end{equation}
where $(x_1,y_1)$ and $(x_2,y_2)$ are any two points on the line but $x_1 \neq
x_2$.  We can write these differences in $x$ as $\Delta x = x_2 - x_1$ and
similarly for $y$, $\Delta y = y_2 - y_1$, which gives for the slope
\begin{equation}
    m = \frac{\Delta y}{\Delta x} .
\end{equation}
If we reduce $\Delta x$, making it smaller and smaller, $\Delta y$ will likewise
shrink since the slope of the line must remain constant.  If we continue to make
$\Delta x$ smaller such that it approaches zero, but never reaches zero, we can
say that $\Delta x$ is infinitesimally small.  This concept of infinitesimals is
fundamental to understanding both derivatives and integrals.

We express this idea of an infinitesimally small quantity as $\dd{x}$.
Mathematically, we would say that in the limit as $\Delta x$ approaches zero,
the result is $\dd{x}$. In math speak, we write this limit as $\lim_{\Delta x \to
0} = \dd{x}$.  So how small is $\dd{x}$?  I can't put an exact number on it.
However, if you choose a value, I can always find a smaller value between your
choice and zero.  As a result of $\dd{x} \to 0$, $\Delta y$ must also tend to
zero, giving us $\dd{y}$.  Remember for a line no matter how small $\dd{x}$ and
$\dd{y}$ become, the ratio of the two will always yield a constant, the slope,
\begin{equation}
    m = \frac{\dd{y}}{\dd{x}} .
\end{equation}

We call the above result a derivative.  In this case, specifically the
derivative of a line.  The derivative of a line is the slope of the line at
every point along it.  So the derivative is the rate in change of the function
($y$) with respect to the independent variable ($x$).  While the derivative of
a function is a constant only if the function is a line, this
concept of measuring the change in a function over a very small change in the
independent variable is kept.

If you are interested, look up the epsilon-delta proof online, it is the formal
proof of what we have just described above.

\subsection{Anatomy of a Derivative}
Above, we showed that the derivative of a line is its slope.  We can extend this
idea of the slope at a point for any function, $f(x)$, more generally.  To
express this idea, we write
\begin{equation}
    \mlbox[
        labelstyle={yshift=0.5em},
        linestyle={}]
        {\dv{}{x}} {differential operator}
    \mlbox[
        labelstyle={below of=m\themathLableNode,below of=m\themathLableNode}]
        {f(x)} {function of $x$}
    = 
    \mlbox[
        labelstyle={xshift=2cm},
        linestyle={out=270,in=90, latex-}]
        {f'(x)} {derivative of $f(x)$} .
\end{equation}
where $f(x)$ is an arbitrary function of $x$, $x$ is the independent variable,
and $f'(x)$ is the derivative of $f(x)$ and also a function of $x$.  The
$\dv{}{x}$ is called the differential operator.  If you are unfamiliar with the
term operator, an operator is a symbol or collection of symbols when taken as a
whole that perform a defined mathematical task.  For example, the `$\dv{}{x}$'
operator indicates that we take the derivative of a function with respect to $x$
just as a `$+$' means that we add the quantities on either side.

\subsection{Practical Differentiation}
We could of course spend the time---as one might in a math course---by deriving
and proving the rules and standard forms for differentiation.  However, since
mathematicians have already done the hard work for us, we can just focus on using them
instead.  All or at least nearly all of the derivatives and rules we will need
for this course can be found in the math appendices under the table of
derivatives.

When I say that all the derivatives that we will need for the course are found
within the table of derivatives, it does not necessarily mean that they are
symbol for symbol the same.  Every equation you will encounter can be rearranged
into one of these common forms.  I've included the table again below for
convenience, but how do we use them?

\subsubsection{Table of Common Derivatives}
\input{tod}

\subsubsection{Using the Derivative Table}
The derivative table above can be used to take the derivatives of many more
functional forms than those shown in the table.  In fact, the table is much more
general than it appears.  The letters `$a$' and `$n$' in the table indicate
constants.  These constants can be anything---a number, a collection of
constants, a function of an independent variable---so long as it isn't a
function of the variable of differentiation.  The letters `$u$' and `$v$' are
both functions of $x$, which can be any arbitrary function so long as it is
differentiable (we discussed this on problem set 1).  But this table doesn't
just work for derivatives with respect to $x$.  They can be with respect to any
independent variable.

So what to do if our problem does not look like one of the derivatives listed on
the table?
\begin{enumerate}
    \item Find a set of functions and constants which transform the original
    function to look like a form in the table; that is, by choosing an appropriate
    function, $u(x)$ and/or $v(x)$, and/or constants, $a$ and/or $n$.

    \item Copy the derivative from the table that matches the transformed
    equation.

    \item Substitute our original constants and functions into the standard
    derivative solution.

    \item Perform the next derivative, repeating the previous three steps until
    $\dv{u}{x} = \dv{x}{x} = 1$ or $\dv{a}{x} = 0$.
\end{enumerate}

One thing you will discover is that for a large majority of these problems, we will
have to take several derivatives.

\begin{example}
    Suppose,
    \begin{equation*}
        f(x) = \frac{1}{(x^2 + c^2)^{1/2}} ,
    \end{equation*}
    where $c$ is a constant.  Find
    \begin{equation*}
        \dv{}{x}\Biggl(\frac{1}
        {(
            \mlbox[
                labelstyle={below of=m\themathLableNode,below of=m\themathLableNode}]
                {x^2 + c^2} {$= u(x)$}
            )
            \mlbox[
                labelstyle={below of=m\themathLableNode,below of=m\themathLableNode}]
                {^{1/2}} {$= n$} 
        }\Biggr) .
    \end{equation*}

    If you look at the table of derivatives, you will not find this exact
    solution listed.  However, looking at the table and our function we can see
    that if we write our function as
    \begin{align*}
        f(x) &= \frac{1}{u^n} , \\
        u &= x^2 + c^2 , \\
        n &= \frac{1}{2} .
    \end{align*}
    The derivative of this function is in our table of derivatives,
    \begin{equation*}
        \dv{}{x}\left( \frac{1}{u^n} \right) = -n \frac{1}{u^{n+1}} \dv{u}{x} .  
    \end{equation*}
    Substituting the original transformed values back in our transform, we find
    \begin{equation*}
        \dv{}{x}\left( \frac{1}{u^n} \right) =
        - \frac{1}{2} \frac{1}{(x^2 + c^2)^{1/2+1}} \dv{}{x} (x^2 + c^2) .
    \end{equation*}
    So you'll see we have another derivative that we need to determine, but at
    least we have simplified our original derivative.  Using the same procedure
    again, we need to transform the derivative term into one in the table.   
    \begin{equation*}
        \dv{}{x} \frac{1}{(x^2 + c^2)^{1/2}} = 
        - \frac{1}{2} \frac{1}{(x^2 + c^2)^{3/2}} \dv{}{x} (
        \mlbox[ ] {x^2} {$= u(x)$}
        + 
        \mlbox[
            labelstyle={below of=m\themathLableNode,below of=m\themathLableNode}]
            {c^2} {$= v(x)$}
        ) .
    \end{equation*}
    Looking at the table of derivatives, we see that we can write our derivative
    in the form
    \begin{equation*}
        \dv{}{x} (u + v) = \dv{u}{x} + \dv{v}{x} .
    \end{equation*}
    Therefore,
    \begin{equation*}
        \dv{}{x} \frac{1}{(x^2 + c^2)^{1/2}} = 
        - \frac{1}{2} \frac{1}{(x^2 + c^2)^{3/2}} \Biggl(\dv{}{x} 
        \mlbox[ ] {x} {$= u(x)$}
        \mlbox[
            labelstyle={below of=m\themathLableNode,below of=m\themathLableNode}]
            {^2} {$= n$}
        + \dv{}{x}
        \mlbox[
            labelstyle={below of=m\themathLableNode,below of=m\themathLableNode}]
            {c^2} {$= a$}
        \Biggr) .
    \end{equation*}
    Now we have two derivatives that we need to evaluate.  For the first, we use
    \begin{align*}
        \dv{}{x} u^n &= n u^{n-1} \dv{u}{x}\\
        &= 2 x^{2-1} \dv{x}{x} .
    \end{align*}
    Since we can see from the table that $\dv{x}{x} = 1$, the result of the
    derivative $\dv{x^2}{x} = 2x$.  Turning our attention to the second derivative,
    we see from the table of derivatives that $\dv{a}{x} = 0$ when $a$ is
    independent of $x$.  Now we've completed all the derivatives we need to
    perform and we can focus on simplifying our expression,
    \begin{align*}
        \dv{}{x} \frac{1}{(x^2 + c^2)^{1/2}} &= - \frac{1}{2} \frac{1}{(x^2 +
        c^2)^{3/2}} (2 x + 0) , \\
        \dv{}{x} \frac{1}{(x^2 + c^2)^{1/2}} &= - \frac{x}{(x^2 +
        c^2)^{3/2}} ,
    \end{align*}
    and we are done.

    This process may have seemed long and laborious, but the more you do them,
    the faster you will become.  Also you'll start to see several steps ahead,
    which will allow you to more rapidly discover the solution.

    We could go one set further and ask what the slope of a function is at a
    particular point, say $x = x_0$.  In this case, we determine the derivative
    and then evaluate our result with $x$ replaced by the value $x_0$.
\end{example}


\subsubsection{Practice Problems}
I will give you some time to try these and then we'll do them together as part
of the tutorial.

\begin{problem}
    Plate Cooling \\
    To find the heat flow from the temperature, we need to take the derivative
    of the temperature with respect to depth; the solution includes the
    following derivative
    \begin{equation*}
        \dv{}{z} \frac{1}{n} \sin \left( \frac{n \pi z}{L} \right)
    \end{equation*}
\end{problem}

\begin{problem}
    Seismic Reflection \\
    To find the minimum travel-time due to a seismic reflection off a boundary,
    we need to take the derivative of t(x) as below,
    \begin{equation*}
        \dv{}{x} v_1^{-1} [ (h^2 + x^2)^{1/2} + (h^2 + (d - x)^2)^{1/2}]
    \end{equation*}
\end{problem}

\begin{problem}
    Decompression Melting \\
    To model partial melting, you need to take the derivative of the melt
    depression term,
    \begin{equation*}
        \dv{}{F} K \left( \frac{X_0}{D + F(1 - D)} \right)^\gamma .
    \end{equation*}
\end{problem}

\begin{problem}
    Gravity on an Ellipsoid \\
    As part of the derivation to determine the gravity of an ellipsoid (remember
    we make an ellipsoidal correction to gravity), we need to differentiate the
    equation of an ellipsoid,
    \begin{equation}
    \dv{}{\theta} \frac{1 - f}{1 - f\sin^2 \theta} .
    \end{equation}
\end{problem}

\subsection{Maxima and Minima}
One of the really useful applications of a derivative is its use to determine
the locations of maxima and minima of a function.  While some functions extend
to $\pm\infty$, there may be individual points where a function decreases in
value before increasing (minima) or increases in value before decreasing
(maxima).  For a maxima, logic tells us that there must be some point along the
path for which the slope on the left is positive and the slope on the right is
negative.  If so, there must be a point where the slope changes sign and,
therefore, must be zero.  So if we want to know where this point occurs, then we
simply find our expression for the derivative and set it equal to zero (i.e.,
$\dv{f(x)}{x} = 0$) and then solve for $x$.  The same must also be true for
minima.  So we may not know, simply by taking the derivative, whether the point
we found is a maximum or minimum, but we can easily check by plotting the
function.

An extension of the maxima and minima is finding the points for which the slope
of a function is maximum or minimum (these points are called inflection points).
How might we find these points using derivatives?

\subsection{Multiple Derivatives}
As alluded to previously, it is possible to take the derivative of a function
multiple times.  The second derivative of a function is quite commonly used to
describe physical systems.  We often call the second derivative the curvature
because it tells us how fast the slope is changing.

To take multiple derivatives, all you need to do is follow the rules to find the
derivative of a function, then take the derivative of the result, and so on
until the desired derivative is reached.  To indicate higher order derivatives,
we include the order of the derivative, e.g., for a second order derivative,
\begin{equation}
    \dv{}{x}\dv{f(x)}{x} = \dv[2]{f(x)}{x} .
\end{equation}
I cannot think of
any derivatives higher than order four in geoscience, and those are uncommon,
used only in flexural problems.  Generally, you won't need
to take more than one or two.

\subsection{Partial Derivatives}
Suppose we had a function that had multiple independent variables, i.e.,
$f(x,y,z)$.  For example, gravity fields due to a sphere, elevation, and
temperatures within the Earth.  We can also take the derivatives of these
functions.  However, when we do so, we specify the independent variable or
multiple independent variables to differentiate with respect to.  We call these
derivatives partial derivatives.  When we take partial derivatives, we change
our notation slightly, exchanging our `$d$' with a `$\partial$'.  Because each
variable is independent of the others, when we take a derivative with respect 



\section{Integrals}
Integrals are the opposite of a derivative, and are called by some an
anti-derivatives.  The fundamental theorem of calculus defines an integral this
way, i.e.,
\begin{equation*}
    f(x) = \dv{}{x} \int_0^x f(x') \dd{x'} ,
\end{equation*}
which basically says the integral is the opposite of a derivative.

We use integrals to determine the area or volume defined by some boundary.  A
submarine uses its velocity and the time it travels in a given direction to
calculate the position of a ship.  The electrical energy you consume is
determined by the integral of your power consumption, which is used to figure
your bill from the power company.  The color of an object is results from the
integral of the spectrum of light reflected off the object and the sensitivity
of the cones in your eyes.

\subsection{Areas}
To perform a derivative, we determine the ratio between small differences in a
function with small differences in the independent variable.  An integral does
the opposite, determining the sum of small areas.  The areas are the products
between small differences of the independent variable and the average of the
function over the same intervals, i.e.,
\begin{equation*}
    F(x) = \sum_{i = 1}^{N} \frac{(f(x_i) + f(x_{i-1}))}{2} (x_i - x_{i-1}) .
\end{equation*}
as the interval $\Delta x_i = x_i - x_{i-1}$ gets smaller and smaller, it becomes
an infinitesimal just as we described previously in the derivative section.  If
we chose our $x_i$'s such that they were all the same size, then over the
interval from $a$ to $b$, we can write
\begin{equation*}
    \dd{x_i} = \dd{x} = \frac{b - a}{N} .
\end{equation*}
As the interval becomes smaller and smaller, i.e., $N \to \infty$, $f(x_i)
\approx f(x_{i-1})$.  This allows us to write
\begin{equation*}
    F = \sum_{i = 1}^{N} f(x_i) \dd{x_i} = \int_a^b f(x) \dd{x} .
\end{equation*}
In fact, the integral symbol, $\int$, was originally an $S$ to signify that an
integral was a sum, so the symbol itself is a stylized $S$ retaining that
connection.

\subsection{Anatomy of an Integral}
Above, we showed that the integral of a line is related to its area beneath a
curve.  We can extend this idea of the area over any interval $[a,b]$ for a function,
$f(x)$, more generally.  To express this idea, we write the definite integral
\begin{equation}
    \mlbox[
        labelstyle={xshift=-2cm, yshift=1em},
        linestyle={out=270,in=180, latex-}]
        {\int_a^b} {integral with limits $[a,b]$}
    \mlbox[
        labelstyle={below of=m\themathLableNode,below of=m\themathLableNode}]
        {f(x)} {integrand, function of $x$}
    \mlbox[ ] {\dd{x}} {variable of integration, $x$}
    = 
    \mlbox[
        labelstyle={xshift=3cm, yshift=1em},
        linestyle={out=270,in=90, latex-}]
        {F(x)} {integral of $f(x)$}
    \mlbox[
        labelstyle={xshift=2cm,below of=m\themathLableNode,below of=m\themathLableNode},
        linestyle={out=90,in=270, latex-}]
        {|_a^b} {evaluated from $a$ to $b$}
    = F(b) - F(a)
\end{equation}
where $f(x)$ is an arbitrary function of $x$, $x$ is the independent variable,
and $F(x)$ is the integral and also a function of $x$.  An integral is often
taken over a bounded region from $[a,b]$.  Not all integrals have limits.  In
these cases, the integral is called an indefinite integral and the solution will
include an arbitrary constant.

\subsection{Practical Integration}
Just as there is for derivatives, a table of integrals can be utilized to
simplify complex integrals.  The steps to using them and find a solution are
similar to those we used for a derivative.  Although, there is an extra step
necessary to transform the infinitesimals from one into another.

\subsubsection{Table of Common Integrals}
\input{toi}

\subsubsection{Using the Integral Table}
We use the integral table in a similar way we use the derivative table.  If our
form is in the table, we copy it and we're done.  But if not, there are multiple
methods for determining the integral to a function.  We will discuss a few here.
This first integration method is by substitution using a transform function
similar to our method for derivatives above,
\begin{enumerate}
    \item Find an equation u(x) which simplifies our function.
    \item Determine $\dv{u}{x}$ and solve for $\dd{u}$ to replace the variable of
    integration.
    \item Copy the integral from the table that matches our transformed
    function.
    \item Substitute our constants and functions into the standard integral solution.
    \item If there are any integrals remaining, repeat steps above or use
    another method.
\end{enumerate}

\begin{example}
    Integrate
    \begin{equation*}
        \int x e^{-x^2} \dd{x} .
    \end{equation*}
    We need a function that can transform our problem above into one that is
    located in the table.  To do so, we look for the integral that has an $\exp$
    term.  It may take a little practice to see which function to use as a
    transform, but in this case, we use $u = -x^2$.
    
    We also need to change our variable of integration to $u$.  To do so, we
    need to start with the derivative of our transform with respect to
    $x$, i.e.,
    \begin{align*}
        \dv{u}{x} &= -2 x \\
        \dd{u} &= -2 x \dd{x} \\
        -\frac{\dd{u}}{2 x} &= \dd{x} .
    \end{align*}
    Now we can transform the integral in to a simpler form,
    \begin{align*}
        \int x e^{-x^2} \dd{x} &= \int x e^u \left(- \frac{\dd{u}}{2 x} \right) \\
        &= \int -\frac{1}{2} e^u \dd{u} .
    \end{align*}
    Solving this integral and substituting back in our original function, we
    find
    \begin{align*}
        -\frac{1}{2} \int e^u \dd{u} &= -\frac{1}{2} e^u \\
        &= -\frac{1}{2} e^{-x^2} .
    \end{align*}

\end{example}

The second integration method requires integration by parts.  To integrate by
parts,
\begin{enumerate}
    \item Find two functions, $u(x)$ and $v(x)$ so that we can write our
    expression as $\int u \dd{v}$ .  We'll need $\dv{v}{x}$ in order to make this
    substitution.
    \item Write the solution to the integration by parts form found in the
    table, i.e., $\int u \dd{v} = u v - \int v \dd{u}$.
    \item Solve the integral $\int v \dd{u}$ by the first or third method.
\end{enumerate}

\begin{example}
    Integrate
    \begin{equation*}
        \int x \cos x \dd{x} .
    \end{equation*}

    Choosing $u = x$ and $v = \sin x$, we can determine $\dd{v}$,
    \begin{align*}
        \dv{v}{x} &= \cos x \\
        \dd{v} &= \cos x \dd{x}
    \end{align*}
    Using the integral solution for parts,
    \begin{equation}
        \int u \dd{v} = u v - \int v \dd{u} ,
    \end{equation}
    substitute in our equations,
    \begin{equation*}
        \int x \cos x \dd{x} = x \sin x - \int \sin x \dd{x} .
    \end{equation*}
    Next, we solve the new integral using the form from the integral table and
    we are done,
    \begin{equation*}
        \int x \cos x \dd{x} = x \sin x + \cos x + C,
    \end{equation*}
    where $C$ is an arbitrary constant as a resulting from the indefinite
    integral.
\end{example}

The third integration method is similar to the substitution method, but
specifically applies to trigonometric substitution.  This method is essentially
a coordinate transformation.  To do this, we apply the following substitutions
\begin{align}
    a^2 - x^2 & x = a \sin \theta \\
    x^2 + a^2 & x = a \tan \theta \\
    x^2 - a^2 & x = a \frac{1}{\cos \theta} .
\end{align}
Once we make our substitution, we follow one of the other two integration
methods above.


\subsubsection{Practice Problems}

\begin{problem}
    Steady-State Temperature \\
    To find the temperature in a half-space, we need to integrate the thermal
    gradient with depth,
    \begin{equation*}
        \int \left( \frac{q_0}{k} - \frac{A}{k} z \right) \dd{z} ,
    \end{equation*}
    where $q_0$, $A$, and $k$ are constants.
\end{problem}

\begin{problem}
    Radioactive Decay \\
    To find the number of parent radionuclides present due to radioactive
    decay, we need to integrate both sides of the following equation,
    \begin{equation*}
        \int_{N_0}^N \frac{\dd{N'}}{N'} = - \int_0^t \lambda \dd{t'} ,
    \end{equation*}
    and once complete, solve for $N$.
\end{problem}

\begin{problem}
    Buried Slab \\
    This integral is similar to one we needed for our buried slab problem,
    \begin{equation*}
        \int_0^R \frac{r}{(r^2 + z^2)^{3/2}} \dd{r} .
    \end{equation*}
\end{problem}

\begin{problem}
    Mass Flux \\
    The mass flux of asthenosphere (mass passing a given point per unit time)
    moved by a tectonic plate can be expressed as
    \begin{equation*}
        \int_0^h \rho u_0 \frac{y}{h} \dd{y} .
    \end{equation*}
\end{problem}

\begin{problem}
    Entropy \\
    In petrology, the entropy is a key thermodynamic parameter.  The change in
    entropy is related to temperature as
    \begin{equation}
        \Delta S = \int_{T_1}^{T_2} \frac{C_P}{T} \dd{T} .
    \end{equation}
\end{problem}


\subsection{Mean of a function}
In addition to finding areas and volumes, we can use integrals to find the mean
of a function.  For example, take a function between two points, say $x_1$ and
$x_2$.  Assuming our function is continuous, logic will tell us that our
function must pass through the mean value at least once on our chosen interval.
This is called the mean value theorem.  To find the mean value of our function
is sort of like taking the arithmetic mean of a discrete set of point and
dividing by the number of points only we take the sum over infinitely small
intervals and divide by the total width of the intervals.  To write this as an
integral,
\begin{equation}
    \overline{f(x)} = \frac{1}{(x_2 - x_1)}\int_{x_1}^{x_2} f(x) \dd{x} ,
\end{equation}
where the bar over the function indicates the mean of the function.

\subsection{Multiple Integrals}
Just as we had for derivatives, we can take multiple integrals.  We might do
this for the accelerations that we measure from an earthquake to determine the
change in displacement of the ground surface during an earthquake.

To determine a multiple integral, we would write
\begin{equation*}
    \int \left(\int f(x) \dd{x}\right) \dd{x} = \iint f(x) \dd{x^2} .
\end{equation*}

If we are taking multiple derivatives, but not for the same coordinate, then we
are finding the integral of a surface or a volume, i.e.,
\begin{align*}
    \text{surface:} \quad & \iint f(x,y) \dd{x} \dd{y} \\
    \text{volume:} \quad & \iiint f(x,y,z) \dd{x} \dd{y} \dd{z}
\end{align*}
Often when we evaluate surface or volume integrals it is advantageous to make a
transform to another coordinate system first in order to exploit symmetries
simplify the integration.  For example, it is easier to integrate the gravity
field due to a sphere in spherical coordinates than in Cartesian coordinates.

\end{document}
