\topic[Workshop Guide]{Workshop Booklet Guide \& Assessment}

\workshopstart

\bigskip

\noindent\textbf{Student Name:} \rule{0.45\textwidth}{0.4pt}
\hfill
\textbf{Student ID:} \rule{0.2\textwidth}{0.4pt}

\bigskip

\section*{Purpose of this booklet}
This booklet is your working record for workshop activities throughout the course. It is intended to:
\begin{itemize}
    \item support your understanding through active participation,
    \item provide a space to develop and record your thinking,
    \item serve as a study resource for tests and exams.
\end{itemize}

You are expected to bring this booklet to every workshop and use it actively during group activities.


\section*{How workshops are assessed}
Workshop performance is assessed using two complementary components:

\subsection*{1. Participation (major component)}
Your participation is assessed during workshops based on your engagement with activities and peers. This includes:
\begin{itemize}
    \item contributing ideas and reasoning during group work,
    \item engaging constructively with others' ideas,
    \item asking questions and working through uncertainty.
\end{itemize}

Participation is recorded throughout the semester. You do not need to speak constantly, but you should be actively involved in your group's work.

\subsection*{2. Booklet (supporting component)}
This booklet provides evidence of your engagement with workshop material.

During the semester, your booklet will be collected \textbf{during tests}  for a quick review. Assessment will focus on:
\begin{itemize}
    \item completeness of activities,
    \item evidence of working, reasoning, and annotations,
    \item overall engagement with the material.
\end{itemize}


\clearpage
\section*{Submission for assessment}
\begin{itemize}
    \item Bring your booklet to each test when requested.
    \item Booklets will be collected at the start of the test and returned at the end where possible.
\end{itemize}

\textbf{If you are unable to submit your booklet at the test:}
\begin{itemize}
    \item You must upload clear photos/scans within 24 hours.
    \item If no submission is received, the booklet will be marked as incomplete for that checkpoint.
\end{itemize}


\section*{Expectations}
This booklet is not marked for perfection or neatness. Instead, it should show:
\begin{itemize}
    \item your attempt to work through problems,
    \item annotations, sketches, and intermediate steps,
    \item corrections and developing ideas.
\end{itemize}

Working collaboratively is expected; however, your booklet should reflect your own engagement with the material.

\section*{Key idea}
Workshops are designed to help you think, discuss, and explore ideas. Your grade reflects both:
\begin{itemize}
    \item how you participate in the learning process, and
    \item the evidence of that engagement captured in this booklet.
\end{itemize}