\weeknum{4}
\topic[Gravity]{Gravity: Density, Structures, and Non-Uniqueness}

\workshopstart

\vspace{-0.75em}
\noindent\textbf{Duration:} 3 hours \hfill \textbf{Setting:} Workshop room (whiteboards; projector)

\section*{Preparation}

\begin{description}
  \item[Pre-reading:]
  \item[Lecture videos:]
  \item[Notes:]
\end{description}

% ---------------------------------------------------------
\section*{Purpose}
This workshop develops intuition about how density varies in porous and compacted materials, how different subsurface structures relate to density contrasts, and how the buried sphere solution illustrates non-uniqueness in gravity. Students work from physical reasoning, analytic solutions, and qualitative cross-section analysis. The session prepares students for Week~5 (filters, upward/downward continuation, reduction-to-the-pole).

% ---------------------------------------------------------
\section*{Learning Outcomes}
\begin{itemize}
  \item Estimate bulk density of porous materials using simple mixing laws.
  \item Explain how compaction and pore-fluid substitution change density.
  \item Predict density contrasts for various geologic structures.
  \item Interpret gravity anomaly width--depth relationships for the buried sphere.
  \item Recognise non-uniqueness and explain why lenticular shallow bodies can mimic deep spheres.
\end{itemize}

% ---------------------------------------------------------
\section*{Roles and Grouping}
Mixed groups of 4--5 with geology and physics/engineering backgrounds.  
Rotate roles: \emph{Facilitator}, \emph{Math Lead}, \emph{Geology Lead}, \emph{Skeptic/QC}, \emph{Recorder}.

% ---------------------------------------------------------
\section*{Session Flow (180 min)}

\timeslot{Warm-up --- ``Why does density matter?''}{10}
In groups, list three factors controlling density of Earth materials.
Share: how do porosity, mineral composition, and pore fluids alter density?
(Students recall Week~2 density work with igneous samples.)

\timelineitem{Density of a Two-Component Porous Sandstone}{35}
\textbf{Goal:} Build intuition for porosity, compaction, and fluid-substitution effects.

\textbf{Tasks (Worksheet A):}
\begin{enumerate}
  \item Derive an expression for the density of a multicomponent system with porosity.
  \item Explore how reducing porosity (compaction from burial) and/or changing the type of fluid filling pores affects bulk density.
  \item Discuss: which fluid contrast would be measurable with gravity alone?
\end{enumerate}

Groups record 3--4 bullet points about physical intuition (e.g., why air-filled rocks are anomalously light; why compaction increases density nonlinearly).

\timelineitem{Geologic Structures to Geologic Anomalies}{30}
\textbf{Goal:} Connect geology to density contrast and anomaly expectations.

\textbf{Structures to analyse (Worksheet B):}

\textbf{Group tasks:}
\begin{enumerate}
  \item Discuss geologic structures that may be described by the simple shapes.
  \item Propose plausible density contrasts ($\Delta\rho$) given typical lithologies.
  \item Predict qualitative anomaly shape: narrow/broad? symmetric/asymmetric? positive/negative?
  \item One group presents each structure; whole class discusses similarities and differences.
\end{enumerate}

\timelineitem{Buried Sphere: Width--Depth and Non-Uniqueness}{30}
\textbf{Goal:} Work with the buried sphere analytic solution and explore non-uniqueness.

\textbf{Given:}
The vertical gravity from a buried sphere of radius $a$, density contrast $\Delta\rho$, and centre depth $z_0$ is:
\[
  g_z(x) = \frac{4\pi G}{3}\Delta\rho\, a^3
            \frac{z_0}{(x^2+z_0^2)^{3/2}}.
\]

\textbf{Tasks (Worksheet C):}
\begin{enumerate}
  \item Measure the \emph{half-width} $x_{1/2}$ from provided plots and derive:
        \[
          \frac{x_{1/2}}{z_0} = \sqrt{2^{2/3} - 1}.
        \]
  \item Explain why increasing depth increases width but decreasing $\Delta\rho$ does not.
  \item Discuss non-uniqueness: why can shallow, lens-shaped bodies (``lenticular'') mimic deep spheres?
  \item Instructor demonstration: Python simulation showing buried sphere vs.\ lenticular bodies with same peak and width.
\end{enumerate}

\timeslot{Superposition: Build-an-Anomaly}{30}
\textbf{Goal:} Construct a plausible local cross-section and its gravity by summing components.\\
\textbf{Provided:} \emph{Superposition Component Library} with scaled kernels (normalized shapes) for: basement step (sedimentary wedge), mafic dike (2D cylinder proxy), felsic pluton (broad polygon), thin regolith cap (negative slab).\\
\textbf{Tasks:}
\begin{enumerate}
  \item Draft a geologically plausible cross-section (2--3 bodies) with $\Delta\rho$ from Activity~A.
  \item Place laterally, set depths/widths, scale amplitudes, and \emph{sum} the responses (trace overlays on graph paper).
  \item Annotate why the geology is plausible (lithology, structure, relative timing).
\end{enumerate}
\textbf{Deliverable:} Labeled cross-section + composite gravity + legend of component contributions.

\timeslot{Match-Three: Cross-Sections \texorpdfstring{$\leftrightarrow$}{↔} Gravity Profiles}{30}
\textbf{Goal:} Practice non-uniqueness reasoning with realistic models.\\
\textbf{Provided:} \emph{Match-Three Handout} --- three local-style cross-sections (A--C) and three gravity profiles (1--3) designed to be confusable.\\
\textbf{Tasks:}
\begin{enumerate}
  \item Match each profile to the most plausible cross-section using: (i) half-width/FWHM (depth cue); (ii) edge slopes/asymmetry (contacts, wedges); (iii) amplitude scaling with $\Delta\rho$ and size.
  \item Propose one additional observation to break any tie (e.g., rock density control, mapping a contact, short-station-spacing gradient, borehole, susceptibility).
  \item Role inversion: \emph{Geologists} present the math-based width--depth rationale; \emph{Phys/Eng} defend the geology.
\end{enumerate}
\textbf{Deliverable:} One page with matches (A$\to$?, B$\to$?, C$\to$?) + tie-breaker justification.

\timeslot{Wrap-up}{15}
\begin{enumerate}
  \item Which factor (porosity, fluid, composition) most strongly affects density in shallow basins?
  \item Which structural density contrast is hardest to distinguish with gravity alone?
\end{enumerate}

% ---------------------------------------------------------
\section*{Materials}
\begin{itemize}
  \item Worksheet A: Two-Component Sandstone Density
  \item Worksheet B: Structure Density Reasoning
  \item Worksheet C: Buried Sphere and Width--Depth
  \item Python code for instructor demo
\end{itemize}

\section*{Assessment}
\begin{itemize}
  \item Worksheet submissions (one per group).
  \item Quality of group discussion and presentations.
  \item Exit ticket reflection.
\end{itemize}

\notepage
\notepage

\subimport{activities/}{activity_sandstone}
\subimport{activities/}{activity_structures}
\subimport{activities/}{activity_buried_sphere}