\begin{activity}{Thermal Refraction by Insulating Sediments}

\section*{Purpose}
To build intuition for how low-conductivity surface layers redistribute
temperature gradients without changing the total heat flow.

\section*{Geological Context}
Consider two continental regions with identical basal heat flow from the mantle:
\begin{itemize}
  \item Region A: crystalline crust exposed at the surface.
  \item Region B: crystalline crust overlain by a thick sedimentary basin.
\end{itemize}

The sediments are dominated by shales and porous clastic rocks with lower
thermal conductivity than basement rock.

\section*{Assumptions}
\begin{itemize}
  \item Steady-state heat flow.
  \item No internal heat production in sediments.
  \item Heat flow is vertical and laterally uniform.
\end{itemize}

\section*{Tasks}
\begin{enumerate}
  \item Using Fourier's law,
  \[
    q = -k \frac{dT}{dz},
  \]
  explain qualitatively how the temperature gradient differs between
  sediments and basement.
  
  \item Sketch temperature versus depth for both regions on the same axes.
  Clearly label:
  \begin{itemize}
    \item sediment thickness,
    \item basement,
    \item relative steepness of gradients.
  \end{itemize}

  \item Indicate where isotherms are compressed and where they are stretched.

  \item Which region would exhibit a higher temperature at the sediment--basement
  interface? Explain why.
\end{enumerate}

\section*{Discussion}
\begin{itemize}
  \item How does the presence of sediments affect measured geothermal gradients?
  \item Which region would you expect to have higher geothermal potential?
  \item Why might sedimentary basins be favorable geothermal targets?
\end{itemize}

\end{activity}
