\begin{activity}{Boundary Conditions and Geological Meaning}

\section*{Purpose}
To understand how boundary conditions encode geological assumptions in numerical
models.

A numerical model does not ``know'' geology—it only knows equations and boundary
conditions.

{\color{red}
\section*{Learning Outcomes}

By the end of this activity, you should be able to:
\begin{itemize}
  \item Distinguish between Dirichlet (fixed value) and Neumann (fixed flux) boundary conditions.
  \item Match appropriate boundary conditions to geological settings (ocean floor, continental lithosphere, cooling intrusion).
  \item Explain how changing a boundary condition changes the physical system the model represents.
  \item Argue why inappropriate boundary conditions can produce physically misleading results even when the numerical solution is mathematically stable.
\end{itemize}
}

\emph{Boundary conditions determine what kind of Earth your model represents.}

\subsection*{Boundary Conditions}

A partial differential equation (PDE) describes how a field varies \emph{inside} a domain. To obtain a unique solution, we must also specify how the domain interacts with its surroundings. This additional information is supplied through \textbf{boundary conditions}.

Consider a potential variable $u(\mathbf{x})$ defined within a domain $\Omega$ bounded by $\partial \Omega$.

\begin{center}
\begin{tikzpicture}[scale=1.1]

\fill[gray!20]
  (0,0) ..
  controls (1.5,1.2) and (3,1.0) ..
  (3.5,0) ..
  controls (4,-1.2) and (2.5,-2.0) ..
  (1,-1.7) ..
  controls (-0.5,-1.2) and (-1.0,-0.5) ..
  cycle;

\draw[thick]
  (0,0) ..
  controls (1.5,1.2) and (3,1.0) ..
  (3.5,0) ..
  controls (4,-1.2) and (2.5,-2.0) ..
  (1,-1.7) ..
  controls (-0.5,-1.2) and (-1.0,-0.5) ..
  cycle;

\node at (1.8,-0.5) {$\Omega$};

\draw[->,thick] (4.4,0.4) -- (3.5,0.2);
\node[right] at (4.5,0.4) {$\partial\Omega$};

% outward normal
\draw[->,thick]
  (2.9,-1.35) -- (3.6,-1.8);

\node[right] at (3.6,-1.8) {$\vu n$};

\end{tikzpicture}
\end{center}

The PDE determines the behaviour of the field within $\Omega$. The boundary conditions describe what happens along $\partial \Omega$.

\subsubsection*{Dirichlet Boundary Condition}

\[
u = f(\mathbf{x})
\qquad \text{on } \partial\Omega
\]

A Dirichlet boundary condition specifies the \emph{value} of the field at the boundary.

\textbf{Physical interpretation:}
The environment outside the model imposes a known value on the boundary.

% Examples:
% \begin{itemize}
% \item Fixed ocean-bottom temperature.
% \item Fixed atmospheric temperature at Earth's surface.
% \item Fixed hydraulic head in a groundwater model.
% \item Fixed gravitational potential on a far-field boundary.
% \end{itemize}

\textbf{What is fixed?}
\[
\boxed{\text{Field value}}
\]

\vspace{0.5em}

\subsubsection*{Neumann Boundary Condition}

\[
\frac{\partial u}{\partial \vu n}
=
g(\mathbf{x})
\qquad \text{on } \partial\Omega
\]

where $n$ is the outward normal direction.

For heat flow, this is often written as

\[
-k\frac{\partial T}{\partial n}=q.
\]

A Neumann boundary condition specifies the \emph{flux} crossing the boundary.

\textbf{Physical interpretation:}
Rather than fixing the field itself, we prescribe the rate at which energy, mass, fluid, or current enters or leaves the domain.

Examples:
\begin{itemize}
\item Mantle heat flux entering the base of the lithosphere.
\item Groundwater recharge entering an aquifer.
\item Prescribed electrical current density in an electrical model.
\end{itemize}

\textbf{What is fixed?}
\[
\boxed{\text{Flux}}
\]

\vspace{0.5em}

\subsubsection*{Robin Boundary Condition}

\[
a(\mathbf{x})u
+
b(\mathbf{x})
\frac{\partial u}{\partial \vu n}
=
c(\mathbf{x})
\qquad \text{on } \partial\Omega
\]

A Robin boundary condition combines a field value and a flux condition.

For example,

\[
-k\frac{\partial T}{\partial n}
=
h(T-T_{\mathrm{atm}})
\]

describes heat exchange between a surface and the atmosphere.

\textbf{Physical interpretation:}
The boundary interacts with an external environment. Neither the field value nor the flux is fixed independently; instead they are linked together.

Examples:
\begin{itemize}
\item Heat exchange between rock and the atmosphere.
\item Cooling of a magmatic intrusion by surrounding country rock.
\item Groundwater leakage through a semi-permeable boundary.
\end{itemize}

\textbf{What is fixed?}
\[
\boxed{\text{A relationship between value and flux}}
\]

\vspace{0.5em}

\subsubsection*{Cauchy Boundary Condition}

\[
u=f(\mathbf{x}),
\qquad
\frac{\partial u}{\partial \vu n}
=
g(\mathbf{x})
\qquad \text{on } \partial\Omega
\]

A Cauchy boundary condition specifies both the field value and its normal derivative.

\textbf{Physical interpretation:}
The boundary is strongly constrained by observations.

Examples:
\begin{itemize}
\item A borehole where both temperature and heat flow are measured.
\item A boundary constrained simultaneously by observed hydraulic head and groundwater flux.
\item Inverse problems where multiple independent observations are available along a boundary.
\end{itemize}

\textbf{What is fixed?}
\[
\boxed{\text{Value and flux}}
\]

\vspace{1em}

\subsubsection*{Summary}

\begin{center}
\begin{tabular}{lll}
\toprule
Boundary Condition & Prescribes & Physical Meaning\\
\midrule
Dirichlet & Field value & Known state\\
Neumann & Flux & Known transfer rate\\
Robin & Value + flux relation & Exchange with surroundings\\
Cauchy & Value and flux & Strong observational constraint\\
\bottomrule
\end{tabular}
\end{center}

\paragraph{Key Idea}
The governing equation describes the physics occurring within the model domain. Boundary conditions describe how the model interacts with the outside world. Two models can use exactly the same PDE and physical properties yet produce completely different results if different boundary conditions are chosen. Choosing boundary conditions is therefore a geological modelling decision, not merely a mathematical one.

\section*{Part A: Grounding in real systems}

Consider the following settings:
\begin{itemize}
    \item Cooling oceanic crust
    \item Stable continental lithosphere
    \item A cooling granite pluton
    \item Antarctic ice sheet geothermal model
\end{itemize}

\textbf{Task:}
For each case, describe what controls the temperature at the boundaries.
\answerbox[4cm]

\section*{Part B: Types of boundary conditions}

Common thermal boundary conditions:
\begin{itemize}
    \item Fixed temperature (Dirichlet)
    \item Fixed heat flux (Neumann)
\end{itemize}

\textbf{Task:}
Match each type of boundary condition to a geological situation.
\answerbox[3cm]

\section*{Part C: Comparing assumptions}

\textbf{Explain} how the model changes if you prescribe:
\begin{itemize}
    \item surface temperature
    \item surface heat flow
\end{itemize}

What is fixed in each case? What is allowed to vary?
\answerbox[3cm]

\section*{Part D: Geological consequences}

\textbf{Compare} appropriate boundary conditions for:
\begin{itemize}
    \item oceanic lithosphere
    \item continental lithosphere
\end{itemize}
\answerbox[3cm]

\section*{Part E: Critical thinking}

\textbf{Task:}
Explain why inappropriate boundary conditions can produce physically misleading
results—even if the numerical solution is mathematically stable.
\answerbox[3cm]

\textbf{Reflection} \\
If two models use the same equation but different boundary conditions, can they
produce completely different results? Why?
\answerbox[3cm]

\bigskip


{\color{red}
\section*{Key Ideas}

Boundary conditions are where geological assumptions enter a numerical model.

\textbf{Boundary conditions define the physical system, not just the mathematics.} Two models with identical equations but different boundary conditions represent completely different geological systems. Choosing a fixed temperature surface is equivalent to assuming an ocean floor in contact with circulating seawater; choosing a fixed heat flux is equivalent to assuming a continental surface where temperatures are free to evolve.

{\color{blue}
\textbf{Dirichlet and Neumann conditions represent different physical constraints.} A Dirichlet condition fixes the value (temperature); a Neumann condition fixes the gradient (heat flux through the surface). The oceanic upper boundary is naturally Dirichlet; the lower lithospheric boundary is often Neumann (fixed mantle heat flux into the base).

\begin{figure}[h]
\centering
\begin{tikzpicture}[scale=1.0,
  box/.style={draw, thick, fill=blue!8, minimum width=3.5cm, minimum height=1.0cm, align=center, font=\small},
  bcbox/.style={draw, thick, rounded corners, minimum width=3.2cm, minimum height=0.8cm, align=center, font=\scriptsize},
  arr/.style={->, thick}
]
  % --- Dirichlet ---
  \node[font=\small\bfseries] at (2.0, 4.2) {Dirichlet (fixed $T$)};
  \draw[fill=cyan!10, draw=cyan!60!black, thick] (0.2, 3.4) rectangle (3.8, 3.9);
  \node[font=\scriptsize, cyan!60!black] at (2.0, 3.65) {$T = T_\text{surf}$ (fixed)};
  \draw[fill=gray!20, thick] (0.2, 0.5) rectangle (3.8, 3.4);
  \node[font=\small, align=center] at (2.0, 1.95) {Lithosphere\\$T(z)$ evolves freely};
  \draw[fill=orange!15, draw=orange!70!black, thick] (0.2, 0.0) rectangle (3.8, 0.5);
  \node[font=\scriptsize, orange!60!black] at (2.0, 0.25) {$q = q_\text{mantle}$ or $T$ fixed};
  \node[font=\scriptsize, align=left] at (2.0,-0.5) {e.g.\ ocean floor: $T_\text{surf} = 0^\circ$C};

  % --- Neumann ---
  \node[font=\small\bfseries] at (7.5, 4.2) {Neumann (fixed $\partial T/\partial z$)};
  \draw[fill=green!10, draw=green!60!black, thick] (5.7, 3.4) rectangle (9.3, 3.9);
  \node[font=\scriptsize, green!60!black] at (7.5, 3.65) {$\partial T/\partial z = $ const (fixed $q$)};
  \draw[fill=gray!20, thick] (5.7, 0.5) rectangle (9.3, 3.4);
  \node[font=\small, align=center] at (7.5, 1.95) {Lithosphere\\$T(z)$ evolves freely};
  \draw[fill=orange!15, draw=orange!70!black, thick] (5.7, 0.0) rectangle (9.3, 0.5);
  \node[font=\scriptsize, orange!60!black] at (7.5, 0.25) {$q = q_\text{mantle}$ (fixed flux in)};
  \node[font=\scriptsize, align=left] at (7.5,-0.5) {e.g.\ continental surface: $q_s$ measured};
\end{tikzpicture}
\caption{Dirichlet (left) fixes temperature at a boundary — appropriate where temperature is controlled externally (ocean floor). Neumann (right) fixes the heat flux gradient — appropriate where a flux is imposed but temperature is free to adjust (base of lithosphere, continental surface heat flow).}
\end{figure}
}

\textbf{Geological understanding is required to choose appropriate boundary conditions.} A numerically stable model with inappropriate boundary conditions can produce results that look reasonable but represent geology that does not exist. Interpreting model output always requires checking that the boundary conditions faithfully represent the geological setting being modelled.
}

\end{activity}