\begin{activity}{Verification Against Analytic Solutions}

\section*{Purpose}
To demonstrate the importance of verifying numerical models against solutions
with known behaviour.

{\color{red}
\section*{Learning Outcomes}

By the end of this activity, you should be able to:
\begin{itemize}
  \item Identify analytic benchmark solutions appropriate for testing a numerical heat flow implementation.
  \item Describe which properties of a numerical solution should be compared with the analytic result (profile shape, heat flow, temporal evolution).
  \item Explain how grid spacing and time-step size affect numerical accuracy and distinguish accuracy from stability.
  \item Justify why verification is essential before using a numerical model for geological prediction.
\end{itemize}
}

Analytic solutions provide benchmarks for testing numerical implementations.

\section*{Task 1: Choose a benchmark}

Identify one analytic solution from earlier in the course that could be reproduced with the finite difference code (e.g., linear steady-state geotherm, half-space cooling, Fourier series plate cooling).

Describe the solution and the conditions it requires.
\answerbox[4cm]

\section*{Task 2: What to compare}

Describe which aspects of the solution you would compare between the analytic result and the numerical output.

Consider: temperature profile shape, heat flow at the surface or base, behaviour at late times.
\answerbox[4cm]

\section*{Task 3: Grid and time-step effects}

Explain how decreasing grid spacing $\Delta z$ and time step $\Delta t$ affects the accuracy of the numerical solution.

What limits how small you can make these in practice?
\answerbox[4cm]

\section*{Task 4: Stable but wrong}

Explain why a numerically \emph{stable} solution is not necessarily \emph{accurate}.

Give an example of a stable but inaccurate result.
\answerbox[3cm]

\section*{Reflection}

Why is verification especially important when modelling complex geological
systems where no field observation can directly confirm the model output?
\answerbox[4cm]

{\color{red}
\section*{Key Ideas}

\textbf{Verification is not the same as validation.} Verification checks that the code correctly solves the equations as written (by comparing to an analytic solution). Validation checks that the equations represent the real system. Both are necessary, and a model can pass verification while still being physically wrong.

\textbf{Convergence with decreasing grid size is the key accuracy test.} If the numerical solution approaches the analytic solution as $\Delta z \to 0$ and $\Delta t \to 0$, the scheme is convergent. The rate at which it converges (first or second order) indicates how quickly errors decrease with refinement.

{\color{blue}
\textbf{Stability and accuracy are independent properties.} A scheme is stable if errors do not grow unboundedly; it is accurate if the truncation error is small. Choosing $\Delta t$ just below the stability limit keeps the solution stable but may leave large truncation errors. Choosing $\Delta t \ll \Delta z^2/2\kappa$ is more accurate but computationally expensive.

\begin{figure}[h]
\centering
\begin{tikzpicture}[scale=1.0,
  quadrant/.style={draw, thick, minimum width=3.0cm, minimum height=1.6cm, align=center, font=\small}
]
  % Quadrant grid
  \draw[thick] (0,0) -- (6,0) -- (6,4) -- (0,4) -- cycle;
  \draw[thick] (3,0) -- (3,4);
  \draw[thick] (0,2) -- (6,2);

  % Axis labels
  \node[font=\small\bfseries, rotate=90] at (-0.6, 2) {Accurate};
  \node[font=\small\bfseries, rotate=90] at (-0.6, 0.8) {Inaccurate (large $\Delta t$)};
  \node[font=\small\bfseries] at (1.5, 4.3) {Unstable};
  \node[font=\small\bfseries] at (4.5, 4.3) {Stable};

  % Quadrant labels + colors
  \fill[red!15]   (0,2) rectangle (3,4);
  \fill[red!15]   (0,0) rectangle (3,2);
  \fill[orange!20](3,2) rectangle (6,4);
  \fill[green!20] (3,0) rectangle (6,2);

  \node[align=center, font=\scriptsize] at (1.5, 3.0) {Accurate\\but unstable\\(blow-up)};
  \node[align=center, font=\scriptsize] at (4.5, 3.0) {Accurate \&\\stable \checkmark\\(small $\Delta t$, $\Delta z$)};
  \node[align=center, font=\scriptsize] at (1.5, 1.0) {Inaccurate\\and unstable\\(avoid)};
  \node[align=center, font=\scriptsize] at (4.5, 1.0) {Stable but\\inaccurate\\ (large $\Delta t$)};
\end{tikzpicture}
\caption{Stability and accuracy are independent. The target is the upper-right quadrant: stable \emph{and} accurate. Stability alone (lower-right) is not sufficient for reliable geological prediction.}
\end{figure}
}
}

\end{activity}
