\documentclass[final,a4paper,12pt]{article}
\usepackage{notes}

\title{Geophysics III}
\author{Steady-State Geotherms}
\date{}

\begin{document}
\maketitle
%\input{../psintro}

\noindent This problem set is worth a total of 20 points to the total problem set score.

\begin{enumerate}
    %\clearpage
    \input{questions/thermal_conductivity_computation}

    %clearpage
    \input{questions/cc_geotherms}

    %clearpage
    \input{questions/radioactive}
\end{enumerate}

\appendix
\section{Computing geotherms}\label{apx:g}

The solution for the temperature at the base of the first layer,
$T_{1}$, can then be estimated using the homogeneous solution,
\begin{equation}
  T_1 = T_0 + \frac{q_0}{k_1} (z_1 - z_0) -
  \frac{A_1}{2 k_1} (z_1 - z_0)^2.
\end{equation}
The heat flow at the base of the first layer is equal to
\begin{equation}
    q_{1} = q_0 - A_1 (z_1 - z_0).
\end{equation}
The homogeneous solution can then be used to estimate the temperature at
the base of the second layer using, $T_1$, as the temperature at the
top and $q_1$ as the heat flow.  The temperature of all subsequent
layers can be found using the same method boot-strapping down through
the domain of interest.  

Solving for the temperature at the base of the $i$-th layer,
\begin{equation}
  T_{i+1} = T_i + \frac{q_i}{k_{i+1}} (z_{i+1} - z_i)
  - \frac{A_{i+1}}{2 k_{i+1}} (z_{i+1} - z_i)^2.
\end{equation}
It is also necessary to know the heat loss at the base of each layer in
order to compute temperature.  Since heat flow is determined by the heat
flowing into a layer plus the heat generated within, the heat flow at
the bottom of each layer is given by
\begin{equation}
    q_{i+1} = q_i - A_{i+1} (z_{i+1} - z_i).
\end{equation}
This solution forms the basis of most 1-D steady-state geotherms.


\section{Thermal isostasy}\label{apx:tis}
It can be shown that the relative elevation difference between a region with
geotherm $T(z)$ and a reference region $T_{\text{ref}}$ is given by
\begin{equation}
    \varepsilon = \alpha \int_{0}^{z_{\text{max}}} [T(z) - T_{\text{ref}}] \,dz.
\end{equation}
For our piecewise geotherms, we can compute an elevation within each depth
interval by
\begin{equation}
    \varepsilon_i = \alpha \int_{z_i}^{z_{i+1}} [T(z) - T_{\text{ref}}] \,dz,
\end{equation}
and summing the individual steps
\begin{equation}
    \varepsilon = \sum_{i = 1}^N \varepsilon_i.
\end{equation}
Within each step, we can integrate difference in temperature by integrating each
geotherm separately and taking the difference since the integral operator is
linear,
\begin{equation}
    \varepsilon_i = \alpha \left[
    \int_{z_i}^{z_{i+1}} T(z) \,dz - \int_{z_i}^{z_{i+1}} T_{\text{ref}} \,dz
    \right].
\end{equation}
The integral of a single geotherm is given by
\begin{align}
    \int_{z_i}^{z_{i+1}} T(z) \,dz & = \int_{z_i}^{z_{i+1}}
        \left[ T_{i} 
        + \frac{q_i}{k_{i+1}} (z - z_i)
        - \frac{A_{i+1}}{2 k_{i+1}} (z - z_i)^2
        \right] \,dz, \\
    & = T_{i} z \left.
        + \frac{q_i}{2 k_{i+1}} (z - z_i)^2
        - \frac{A_{i+1}}{6 k_{i+1}} (z - z_i)^3
        \right|_{z_i}^{z_{i+1}}, \\
    & = T_{i} (z_{i+1} - z_i) 
        + \frac{q_i}{2 k_{i+1}} (z_{i+1} - z_i)^2
        - \frac{A_{i+1}}{6 k_{i+1}} (z_{i+1} - z_i)^3 .
\end{align}

\end{document}