\documentclass[11pt,a4paper]{article}
\usepackage{notes}
\usepackage{siunitx}
\usepackage{fancyhdr}
\sisetup{detect-all=true}

\pagestyle{fancy}
\fancyhf{}
\lhead{Geophysics IIIA: Potential Fields \& Geothermics}
\rhead{Week 3 Workshop}
\cfoot{\thepage}

\title{Week 3 -- Exploring \& Understanding Inversion}
\author{Geophysics 3A: Potential Fields \& Geothermics}
\date{\today}

\begin{document}
\maketitle

\vspace{-0.75em}
\noindent\textbf{Duration:} 3 hours \hfill 
\textbf{Setting:} Lecture room with laptops (sphere inversion app)

% ----------------------------------------------------------
\section*{Purpose}
This workshop introduces the structure of inverse problems in potential-field geophysics, contrasting forward prediction with inverse estimation. Students explore the role of misfit, data weighting, and regularization, and examine how non-linearity and non-uniqueness arise in gravity modelling. Through a sequence of activities, they classify linear vs.\ non-linear models, reason qualitatively about non-uniqueness, and engage in hands-on inversion of a buried sphere using grid search and Gauss--Newton methods.

% ----------------------------------------------------------
\section*{Learning Outcomes}
\begin{itemize}
  \item Distinguish forward and inverse problems and articulate the role of misfit, weighting, and regularization.
  \item Classify simple geophysical relations as linear or non-linear and propose appropriate linearizations.
  \item Recognize manifestations of non-uniqueness in gravity and magnetic profiles and propose discriminating data.
  \item Derive or confirm analytical Jacobian expressions for the buried sphere model.
  \item Interpret misfit surfaces, inversion paths, and the effect of noise and damping.
  \item Explain parameter trade-offs and resolution limitations inherent to the buried-sphere problem.
\end{itemize}

\textit{Reference concepts align with Week 3 notes on forward/inverse modelling, misfit structure, and non-linear least squares.}

% ----------------------------------------------------------
\section*{Roles and Grouping}
Create mixed groups (4--5 students). Recommended roles: \emph{Facilitator}, \emph{Analyst}, \emph{Recorder}, \emph{Skeptic}, \emph{Reporter}. Rotate roles between major activities.

% ----------------------------------------------------------
\section*{Session Flow (180 min)}

% ---------- 0--30 ----------
\subsection*{0--30 min: Mini-lecture --- Forward vs Inverse, Misfit, Weighting, Challenges}
\begin{itemize}
  \item Forward vs.\ inverse models: $m \mapsto g(m)$ vs.\ $d \mapsto m$.
  \item Misfit: residual $r=d-g(m)$; weighted least squares $\Phi_d=\lVert W_d r\rVert_2^2$.
  \item Role of $W_d$ (noise) and $W_m$ (model constraints); damping $\lambda$.
  \item Challenges: non-linearity (Jacobian needed), non-uniqueness (multiple $m$ give similar $g(m)$), ill-conditioning.
\end{itemize}

% ---------- 30--60 ----------
\subsection*{30--60 min: Activity --- Linear or Non-linear?}
Using the ten-equation handout:
\begin{itemize}
  \item Identify model parameters and classify each relation as \emph{linear}, \emph{non-linear}, or \emph{linearizable}.
  \item If non-linear, propose a linearization (log-transform, re-parameterization, Taylor expansion/Jacobian).
  \item Deliverable: brief notes per group.
\end{itemize}

% ---------- 70--90 ----------
\subsection*{70--90 min: Activity D --- Non-uniqueness (Jigsaw)}
Curves-only reasoning:
\begin{itemize}
  \item Scenario A: gravity profiles --- two distinct geometric models can reproduce similar anomalies.
  \item Scenario B: magnetic dike profiles --- width/depth/magnetization trade-offs.
  \item For \emph{each} curve: propose at least two plausible geologic/geometric models.
  \item List discriminating data (station density, gradients, susceptibility/remanence, multi-component data, mapping).
  \item Deliverable: 1-paragraph argument explaining which model to pursue first and what data to collect next.
\end{itemize}

% ---------- 90--100 ----------
\subsection*{90--100 min: Buried Sphere --- Introduction}
\begin{itemize}
  \item App overview: Data \& Model view; Misfit view; controls for acquisition, noise, synthetic model.
  \item Construct a true model \& generate data (freeze a single noise realization).
\end{itemize}

% ---------- 100--120 ----------
\subsection*{100--120 min: Buried Sphere --- Jacobian Derivatives}
\begin{enumerate}
  \item Students derive/confirm analytical expressions for 
    $\partial g/\partial z$, 
    $\partial g/\partial R$, 
    $\partial g/\partial \Delta\rho$.
  \item Compare with the app's Jacobian; discuss relative sensitivities: 
    amplitude $\propto R^3\Delta\rho$, width controlled by depth $z$.
\end{enumerate}

% ---------- 120--155 ----------
\subsection*{120--155 min: Buried Sphere --- Investigation}
\begin{itemize}
  \item \textbf{Grid search:} Explore pairwise misfit slices and 3-D misfit surface. Identify local vs global minima.
  \item \textbf{Inversion path:} Run Gauss--Newton with different $\lambda$; examine step length and stability.
  \item \textbf{Effect of noise:} Increase $\sigma$; observe widening misfit basins and parameter uncertainty.
  \item Discuss links to curvature, conditioning, and resolution.
\end{itemize}

% ---------- 155--170 ----------
\subsection*{155--170 min: Buried Sphere --- Non-uniqueness}
\begin{itemize}
  \item Demonstrate parameter trade-offs (e.g.\ $R^3\Delta\rho$) and resolution ellipses.
  \item Compare inversion results with misfit contours; highlight underdetermined directions.
\end{itemize}

% ---------- 170--180 ----------
\subsection*{170--180 min: Wrap-up}
\begin{itemize}
  \item Key takeaways: forward/inverse distinctions, misfit and weighting, non-linearity, non-uniqueness, damping.
  \item Open Q\&A; preview of next week's material.
\end{itemize}

% ----------------------------------------------------------
\section*{Materials}
\begin{itemize}
  \item Linear/non-linear handout or slide deck.
  \item Activity D handouts (gravity and magnetic curves-only plots).
  \item Buried sphere app (laptop per group).
  \item Whiteboard or shared notepad for Jacobian derivations.
\end{itemize}

\section*{Assessment}
\begin{itemize}
  \item Activity notes (classification \& argumentation).
  \item Jacobian derivation steps.
  \item Short reflection or exit-ticket on non-linearity and non-uniqueness.
\end{itemize}

\end{document}
\section*{References (suggested for students)}
\begin{thebibliography}{9}
\bibitem{Blakely} R. Blakely, \textit{Potential Theory in Gravity and Magnetic Applications}, Cambridge University Press, 1995.
\bibitem{Telford} W. M. Telford, L. P. Geldart, and R. E. Sheriff, \textit{Applied Geophysics}, 2nd ed., Cambridge University Press, 1990.
\bibitem{Menke} W. Menke, \textit{Geophysical Data Analysis: Discrete Inverse Theory}, 3rd ed., Academic Press, 2012.
\bibitem{Aster} R. C. Aster, B. Borchers, and C. H. Thurber, \textit{Parameter Estimation and Inverse Problems}, 3rd ed., Elsevier, 2019.
\bibitem{Tarantola} A. Tarantola, \textit{Inverse Problem Theory and Methods for Model Parameter Estimation}, SIAM, 2005.
\end{thebibliography}

\end{document}
