\begin{activity}{Half-Space Cooling of the Oceanic Lithosphere}

\section*{Purpose}
To develop physical intuition for transient cooling of the oceanic lithosphere
using the half-space cooling model.

{\color{red}
\section*{Learning Outcomes}

By the end of this activity, you should be able to:
\begin{itemize}
  \item State the assumptions of the half-space cooling model and identify which geological settings they approximate.
  \item Sketch temperature--depth profiles for increasing seafloor ages and identify the thermal boundary layer.
  \item Derive or explain the $\sqrt{t}$ dependence of thermal boundary layer thickness and heat flow.
  \item Explain the physical reason why the half-space model predicts unbounded lithospheric thickening.
\end{itemize}
}

\section*{Model Assumptions}
\begin{itemize}
  \item The mantle is initially at uniform temperature.
  \item Surface temperature is suddenly reduced and held fixed.
  \item No fixed lithospheric thickness is imposed.
\end{itemize}

\section*{Solution Form}
The temperature evolution is given by
\[
T(z,t) = T_m \,\mathrm{erf}\!\left(
\frac{z}{2\sqrt{\kappa t}}
\right).
\]

\section*{Tasks}
\begin{enumerate}
  \item Sketch temperature--depth profiles for increasing times (e.g., 10, 50, 100~Ma). Label the thermal boundary layer.
  \answerbox[6cm]

  \item Identify the thermal boundary layer and describe how its thickness evolves.
  \answerbox[2cm]

  \item Estimate how the depth to a given isotherm scales with seafloor age.
  \answerbox[3cm]
\end{enumerate}

\section*{Geological Interpretation}
\begin{itemize}
  \item Oceanic lithosphere thickens thermally with age.
  \item Heat flow decreases as the plate cools away from mid-ocean ridges.
\end{itemize}

Explain why the half-space cooling model predicts unbounded lithospheric thickening.
\answerbox[3cm]

{\color{red}
\section*{Key Ideas}

{\color{blue}
\textbf{Oceanic lithosphere is a thermal boundary layer.} The lithosphere defined thermally is the region of the mantle that has cooled significantly below the ambient mantle temperature. Its thickness grows as $\ell_T \sim 2\sqrt{\kappa t}$ and is proportional to the square root of seafloor age.

\begin{figure}[h]
\centering
\begin{tikzpicture}[scale=0.9]
  % T-z profiles at increasing ages
  \draw[->] (0, 0) -- (5.5, 0) node[right] {$T/T_m$};
  \draw[->] (0, 0) -- (0,-5.5) node[below] {$z$ (km)};

  % erf profiles at different ages
  % Age 10 Ma: shallow TBL
  \draw[thick, blue!40, domain=0:1.5, samples=30]
    plot ({1.0*(1 - exp(-\x*\x*2.5))}, {-\x*30});
  \draw[thick, blue!40] ({1.0*(1-exp(-1.5*1.5*2.5))}, -1.5*30) -- (1.0, -5.0);
  \node[blue!60, font=\scriptsize, right] at (0.8,-0.5) {10 Ma};

  % Age 50 Ma: deeper TBL
  \draw[thick, blue!70, domain=0:3.0, samples=40]
    plot ({1.0*(1-exp(-\x*\x*0.5))}, {-\x*13.3});
  \node[blue!70, font=\scriptsize, right] at (0.5,-2.0) {50 Ma};

  % Age 100 Ma: deeper still
  \draw[thick, blue, domain=0:5.0, samples=50]
    plot ({1.0*(1-exp(-\x*\x*0.25))}, {-\x*9.4});
  \node[blue, font=\scriptsize, right] at (0.3,-3.5) {100 Ma};

  % TBL depth markers
  \draw[dashed, gray] (0,-1.2) -- (-0.5,-1.2) node[left, font=\scriptsize] {$\ell_T(10)$};
  \draw[dashed, gray] (0,-2.7) -- (-0.5,-2.7) node[left, font=\scriptsize] {$\ell_T(50)$};
  \draw[dashed, gray] (0,-4.0) -- (-0.5,-4.0) node[left, font=\scriptsize] {$\ell_T(100)$};

  % sqrt-t annotation
  \draw[->, thick, orange] (3.5,-1.2) -- (3.5,-4.0)
    node[midway, right, font=\small] {$\propto\sqrt{t}$};
\end{tikzpicture}
\caption{Half-space cooling: the thermal boundary layer thickens as $\sqrt{t}$ (orange arrow). At 10, 50 and 100 Ma the temperature--depth profiles deepen, converging toward the mantle temperature $T_m$ at greater depth with increasing age.}
\end{figure}
}

\textbf{Heat flow decreases as $t^{-1/2}$.} As the lithosphere thickens with age, the surface temperature gradient decreases and surface heat flow falls proportionally. This explains the systematic decrease in heat flow with increasing distance from mid-ocean ridges.

\textbf{The model has no basal limit.} Because there is no imposed base to the half-space, the thermal boundary layer grows indefinitely. In reality, the observed flattening of bathymetry and heat flow for old (${>}80$~Ma) plates requires a fixed lower boundary, which is captured by the plate cooling model.
}

\end{activity}