\begin{activity}{Thermal vs Chemical Diffusion}

\section*{Purpose}
To compare transient thermal diffusion with chemical diffusion processes familiar from igneous and metamorphic petrology.

{\color{red}
\section*{Learning Outcomes}

By the end of this activity, you should be able to:
\begin{itemize}
  \item Write the general diffusion equation and identify the diffusivity that controls the rate of smoothing.
  \item Explain why diffusion processes (thermal and chemical) smooth spatial gradients over time.
  \item Compare thermal and chemical diffusivities and explain why chemical gradients can be preserved long after thermal gradients are erased.
  \item Describe how multi-component diffusion (where species diffuse at different rates) differs from single-field thermal diffusion.
\end{itemize}
}

\section*{Diffusion Framework}
Both thermal and chemical diffusion can be written in the general form
\[
\frac{\partial \phi}{\partial t} = D \nabla^2 \phi,
\]
where $\phi$ represents temperature or chemical concentration, and $D$ is a diffusivity.

\section*{Tasks}
\begin{enumerate}
  \item Explain why both thermal and chemical diffusion act to smooth spatial gradients over time.
  \answerbox[3cm]

  \item Chemical diffusion in minerals often preserves sharp zoning. Give two reasons why sharp compositional gradients may survive despite diffusion.
  \answerbox[3cm]

  \item Describe how multi-component diffusion differs from thermal diffusion involving a single scalar field.
  \answerbox[3cm]
\end{enumerate}

\section*{Comparison}
\begin{itemize}
  \item Thermal diffusion: single field, relatively uniform diffusivity.
  \item Chemical diffusion: multi-component, potentially rate-limited by the slowest species.
\end{itemize}

\section*{Reflection}
Why is heat generally a poorer long-term recorder of geological processes than composition?

\answerbox[3cm]

{\color{red}
\section*{Key Ideas}

\textbf{Diffusion smooths gradients.} Both thermal and chemical diffusion act to eliminate spatial variations in the field (temperature or concentration). Steep gradients drive faster smoothing; uniform fields are stable.

\textbf{Thermal diffusivity is much larger than most chemical diffusivities.} Heat diffuses through rock roughly $10^4$--$10^6$ times faster than most major elements. As a result, temperature gradients in the crust equilibrate on timescales of millions of years, whereas chemical zoning in minerals can survive for billions of years.

\textbf{Multi-component chemical diffusion can be rate-limited.} When multiple chemical species must diffuse simultaneously, the slowest-diffusing component controls the overall rate. This is why garnet zoning records precise closure temperatures even in terranes that have experienced extensive heating.
}

\end{activity}