\weeknum{9}
\topic[Transient Heat Flow]{Geothermics: Transient Heat Flow}

\workshopstart

\vspace{-0.75em}
\noindent\textbf{Duration:} 3 hours \hfill \textbf{Setting:} Workshop room

\section*{Preparation}

\begin{description}
  \item[Pre-reading:] Course Notes, Ch. 11
  \item[Lecture videos:]
  \item[Notes:]
\end{description}

\section*{Purpose}

This workshop develops intuition for transient heat flow: how thermal diffusivity sets the timescales and length scales over which thermal signals propagate, how half-space and plate cooling models describe the thermal evolution of oceanic lithosphere, and how latent heat modifies cooling in Stefan problems.

\section*{Learning Outcomes}
Students will:
\begin{itemize}
  \item Understand how thermal diffusivity controls timescales.
  \item Compare thermal and chemical diffusion processes.
  \item Interpret half-space and plate cooling models.
  \item Use Fourier series to conceptualize thermal evolution.
\end{itemize}

\section*{Session Flow (\timelinelength\ min)}

% {\color{red}
% \timeline[slot]{Warm-up --- How long does it take to cook a roast?}{10}

% Ask students: \emph{Why does a larger roast take longer to cook in the oven even at the same temperature? What does this tell us about heat moving through rock?} Elicit the idea that the time for heat to penetrate scales with the \emph{square} of the distance. Connect to the diffusion equation and introduce thermal diffusivity $\kappa$ as the quantity that controls this scaling.
% }

\timeline{Warm-up: Wedge vs.\ Orogen}{10}

Both accretionary wedges and orogens form at convergent margins, but they experience very different thermal histories. Before we build a detailed model for one of them later today, predict how surface heat flow evolves over the life of each.

\subsection*{Accretionary wedge}
Sediment and crust are continuously scraped off a subducting oceanic plate and stacked at the trench. The wedge is built from material that has spent little time near the surface -- it arrives cold, wet, and is rapidly buried by whatever is scraped off next.

\begin{itemize}
  \item While subduction is active, is heat flow through the wedge likely to be higher or lower than a stable craton? What is actually supplying heat to this material, and what is working against it staying warm?
  \item Once subduction stops -- no more cold material arriving, no more rapid burial -- what happens to heat flow? Does it recover to the same level as before subduction started, or could it overshoot?
\end{itemize}

\subsection*{Orogen}
Continental collision thickens the crust -- doubling it is not unusual. The thickened crust contains more total radiogenic material than before, buried deeper than it used to be.

\begin{itemize}
  \item Right after thickening, is the crust hotter or colder than it will eventually become? (Think about how long it takes buried heat production to actually raise temperatures at depth -- you'll compute this properly in Activity 9.A.)
  \item Once the orogen has had a long time to adjust, how does its eventual heat flow compare to a stable craton -- higher, lower, or about the same?
\end{itemize}

\subsection*{Sketch both}
On the axes below, sketch your predicted heat flow through time for \emph{both} settings, on the same plot. The dashed line marks typical cratonic heat flow, for reference.

\begin{center}
\begin{tikzpicture}[x=1cm, y=1cm]
  \draw[thick,->] (0,0) -- (9.5,0) node[right] {time};
  \draw[thick,->] (0,0) -- (0,6.5) node[above] {heat flow};
  \draw[dashed, gray] (0,2) -- (9,2);
  \node[gray, right, font=\scriptsize] at (9,2) {craton};
\end{tikzpicture}
\end{center}

\noindent\textit{We will not resolve the wedge case in detail today -- keep your prediction in mind as a point of contrast. The orogen case is exactly what Activity 9.B (Collisional Thermal Evolution) works through properly, including why the answer to the first orogen question is more surprising than it looks.}


\timeline[slot]{Mini-lecture --- The diffusion equation and thermal diffusivity}{15}

\begin{itemize}
  \item The heat (diffusion) equation: $\partial T/\partial t = \kappa\, \nabla^2 T$.
  \item Thermal diffusivity $\kappa = k/(\rho c_p)$: what controls it, typical values for crustal rocks.
  \item Scaling argument: thermal length $z \sim \sqrt{\kappa t}$; thermal timescale $t \sim z^2/\kappa$.
  \item Key difference from steady state: temperature \emph{evolves} in time.
\end{itemize}

\timeline[item]{Thermal Diffusivity, Thermal Length, and Timescales}{45}
\begin{itemize}
  \item Review $\kappa = k / (\rho c_p)$.
  \item Emphasize modest variation in crustal rocks.
  \item Contrast sediments, rock, and ice.
  \item Penetration depth scaling: $z \sim \sqrt{\kappa t}$.
  \item Borehole paleothermometry vs ice-core records.
  \item Depth--resolution tradeoffs.
\end{itemize}

\timeline[item]{Thermal vs Chemical Diffusion}{30}
\begin{itemize}
  \item Both obey $\sqrt{Dt}$ scaling.
  \item Chemical diffusion often multi-component.
  \item Zoning persistence vs thermal smoothing.
\end{itemize}

\timeline[slot]{Break}{10}

\timeline[item]{Half-Space Cooling}{35}
\begin{itemize}
  \item Assumptions and boundary conditions.
  \item Growth of thermal boundary layer.
  \item Oceanic lithosphere as a thermal clock.
\end{itemize}

\timeline[item]{Plate Cooling and Fourier Series}{30}
\begin{itemize}
  \item Fixed plate thickness as boundary condition.
  \item Fourier series as superposition of thermal modes.
  \item Mode decay times and geological interpretation.
\end{itemize}

\timeline[item]{Stefan Problems in Geology}{30}
\begin{itemize}
  \item Latent heat as effective heat capacity.
  \item Cooling times of dikes vs sills.
  \item When analytic solutions break down.
\end{itemize}

{\color{red}
\timeline[slot]{Wrap-up}{10}

Bring the session together. Ask: \emph{What single quantity controls how fast heat moves through rock, and what are its typical values?} Revisit the timescale scaling $t \sim z^2/\kappa$ and ask students to give one geological example from each activity where this scaling is evident. Highlight the progression from the simple half-space to the plate model to the Stefan problem as increasing physical realism.
}

{\color{red}
\timeline[slot]{Preview: Finite Differences}{10}

Analytic solutions to the diffusion equation require simplifying assumptions: constant properties, simple geometries, and idealized boundary conditions. Real geological systems violate these. Next week we develop finite difference methods, which replace the differential equation with a local energy balance that can be applied to arbitrarily complex geometries and boundary conditions. You will first derive the finite difference equations from first principles and then solve the heat equation yourself by hand---becoming the computer.
}

\workshopend
\notepage
\notepage

\subimport{activities/}{activity_collisions}
\subimport{activities/}{activity_thermal_timescales}
\subimport{activities/}{activity_oceanic_cooling}