\documentclass[11pt,a4paper]{article}
\usepackage{notes}
\usepackage{fancyhdr}
\author{Geothermics: Finite Differences}
\date{\today}

\pagestyle{fancy}
\fancyhf{}
\lhead{Geophysics 3A: Potential Fields \& Geothermics}
\rhead{Week 10 Workshop}
\cfoot{\thepage}

\title{Week 10 -- Workshop Plan\\Finite Differences}
\author{Geophysics 3A: Potential Fields \& Geothermics}
\date{\today}

\begin{document}
\maketitle

\section*{Workshop Overview}

This workshop builds intuition for finite difference methods through physical reasoning and interactive activities. Students first derive the heat equation from energy conservation, then experience how numerical models work by acting as nodes that update temperature based on their neighbours.  The workshop then explores how boundary conditions represent geological assumptions and how numerical stability reflects limits on how heat can physically propagate.  By the end, students should understand that finite difference methods are simply local energy balance, that heat flow emerges from neighbour interactions, and that both boundary conditions and stability have direct physical meaning.

\section*{Learning Objectives}
Students will:
\begin{itemize}
  \item Understand finite difference equations as local statements of energy conservation.
  \item Develop intuition for how numerical solutions evolve through neighbour interactions.
  \item Interpret boundary conditions as geological assumptions.
  \item Understand numerical stability as a limit on physical information transfer.
\end{itemize}

\section*{0:00--0:30 \quad From Energy Conservation to Finite Differences}

\textbf{Concept} \\
Finite difference equations arise from applying conservation of energy to small
rock volumes.

\begin{itemize}
  \item Nodes represent local rock volumes.
  \item Heat flux enters and leaves each node.
  \item Heat production adds energy locally.
\end{itemize}

\textbf{Activity:} From Conservation of Energy to Finite Differences

The equation is \emph{not a numerical trick}.  The Laplacian as a measure of curvature (difference from neighbours).  This is just bookkeeping of energy.

\section*{0:30--1:05 \quad Finite Differences as a Physical Process}

\textbf{Concept} \\
Numerical models evolve through local interactions between neighbouring nodes.

\textbf{Activity:} Human Finite Difference Solver

\textbf{Discuss:}
\begin{itemize}
  \item smoothing of temperature
  \item spreading of heat
  \item role of gradients
\end{itemize}

\section*{1:05--1:25 \quad Boundary Conditions as Geological Assumptions}

\textbf{Concept:} Boundary conditions determine what physical system the model represents.
\begin{itemize}
  \item Fixed temperature (Dirichlet)
  \item Fixed heat flux (Neumann)
\end{itemize}

\textbf{Activity:} Boundary Conditions and Geological Meaning

\section*{1:25--1:35 \quad Break}

\section*{1:35--2:00 \quad When Analytic Solutions Break Down}

\textbf{Concept}
\begin{itemize}
  \item Analytic solutions require simplifications:
  \begin{itemize}
      \item constant properties
      \item simple geometries
      \item simple boundary conditions
  \end{itemize}
  \item Real geology violates these assumptions.
\end{itemize}

\textbf{Discussion prompts}
\begin{itemize}
  \item What makes geological systems difficult to solve analytically?
  \item Why do irregular boundaries force numerical approaches?
\end{itemize}

Your Iberia exercise on the problem has several issues that make it impossible to develop an analytic form, including variable sediment thickness, lateral structure, varying boundary condition.

\section*{2:00--2:40 \quad Numerical Stability and Information Transfer}

\textbf{Concept} Stability limits impose a ``speed limit'' on thermal information:
\[
\Delta t < \frac{\Delta x^2}{2\kappa}
\]

\begin{itemize}
  \item Heat propagates gradually through diffusion.
  \item Numerical schemes must respect this.
\end{itemize}

\textbf{Activity:} Stability, Time Stepping, and Information Transfer
\begin{itemize}
    \item small vs large time steps
    \item what happens when updates are too large
    \item stability = causality
    \item information cannot skip nodes
\end{itemize}

Instability is not just numerical---it is physically impossible behaviour.

\section*{2:40--3:00 \quad Wrap-Up and Synthesis}

\textbf{Key connections}

\begin{itemize}
  \item Finite differences = local energy balance + neighbour interaction
  \item Boundary conditions = geological assumptions
  \item Stability = physical constraint on information transfer
\end{itemize}

\textbf{Discussion}

\begin{itemize}
  \item How does the numerical method compare to analytic solutions?
  \item What advantages do numerical models provide?
  \item Where might these methods be applied beyond heat flow?
\end{itemize}

\textbf{Preview}

\begin{itemize}
  \item Extension to 2-D systems
  \item Application to real datasets (Iberia Abyssal Plain)
\end{itemize}

\end{document}