%\usetikzlibrary{decorations.pathmorphing, arrows.meta}
%\usepackage{multicol}
%\usepackage{pgfplots}

\weeknum{5}
\topic[Isostasy]{Isostasy: Ocean--Continent Structure \& Isostatic Gravity}

\workshopstart

\vspace{-0.75em}
\noindent\textbf{Duration:} 3 hours \hfill \textbf{Setting:} Workshop room (small-group whiteboard work)

\section*{Preparation}

\begin{description}
  \item[Pre-reading:]
  \item[Lecture videos:]
  \item[Notes:]
\end{description}

\section*{Purpose}

This workshop develops a quantitative understanding of isostatic equilibrium and its gravity expression by combining column-based reasoning, ocean--continent structure, and slab gravity approximations. Airy and Pratt isostasy are treated as end-member models, with emphasis on why real Earth structure requires a mixed or layered approach involving the lithospheric mantle.

\section*{Learning Outcomes}

By the end of this workshop, you should be able to:

\begin{itemize}
    \item Explain isostatic equilibrium as equal lithostatic pressure (or mass per unit area) at a compensation depth.
    \item Derive and interpret Airy and Pratt isostasy as conceptual end members.
    \item Explain why crust-only isostatic models fail for ocean--continent systems.
    \item Derive the Airy isostatic gravity correction using slab approximations.
    \item Explain why gravity anomalies over high topography may be small.
\end{itemize}

\section*{Session Flow (180 min)}

\timeslot{Warm-Up: Defining Isostasy}{10}

Individually, write a one-sentence definition of isostatic equilibrium. These definitions are compared in pairs and then within your group, with the goal of converging on a precise statement.

\smallskip
\noindent
\textbf{Target idea:}
\begin{quote}
Isostatic equilibrium requires equal lithostatic pressure (or mass per unit area) at a compensation depth beneath adjacent columns.
\end{quote}

Groups assign roles (facilitator, skeptic, recorder, assumption checker).

\timeslot{Mini-Lecture: Airy, Pratt, and Mixed Isostasy}{20}

A short instructor-led lecture reviews:

\begin{itemize}
    \item Airy isostasy: variable crustal thickness, uniform density
    \item Pratt isostasy: constant thickness, lateral density variations
    \item Why neither end member fully describes the Earth
    \item The role of lithospheric mantle and partial compensation
\end{itemize}

This provides the conceptual framework for the two analytical activities that follow.

\timelineitem{Oceanic vs Continental Columns}{40}

\subsection*{Purpose}

To test whether crust-only Airy isostasy can account for observed ocean--continent elevation differences, and to explore how including the lithospheric mantle resolves inconsistencies.

\subsection*{Crust-Only Airy Isostasy}

Each group balances an oceanic and a continental column under the assumption that compensation occurs solely through crustal thickness. You will:

\begin{itemize}
    \item Enforce equal mass per unit area above a compensation depth.
    \item Solve for the oceanic crust density required for equilibrium.
    \item Compare your result with realistic oceanic crust densities.
\end{itemize}

You should find that the required density is unrealistic, indicating a missing component in the model.

\subsection*{Two-Layer Lithosphere Extension}

The model is then extended to include a lithospheric mantle beneath both oceanic and continental crust. You will:

\begin{itemize}
    \item Rebalance the columns including mantle lithosphere thickness differences.
    \item Estimate the density contrast between oceanic and continental lithospheric mantle.
    \item Relate this contrast to mineralogical differences (e.g.\ garnet content).
\end{itemize}

\timeslot{Whole-Class Synthesis}{10}

A brief instructor-led discussion synthesises results from Activity 1, focusing on:

\begin{itemize}
    \item Why crust-only Airy isostasy fails
    \item Why the lithospheric mantle must be included
    \item Where Pratt-like behaviour may arise
\end{itemize}

\timelineitem{Isostatic Gravity Correction}{40}

\subsection*{Purpose}

To derive the Airy isostatic gravity correction and explain why large topography does not necessarily produce large gravity anomalies.

\subsection*{Airy Mass Balance}

You will show that, under Airy isostasy,
\[
\rho_c h = (\rho_m - \rho_c)\, r ,
\]
and derive an expression for the crustal root thickness $r(x)$.

\subsection*{Infinite Slab Gravity}

Using the slab approximation,
\[
\Delta g = 2\pi G \rho t ,
\]
you will compute the gravitational effects of surface topography and of the compensating crustal root.

\subsection*{Cancellation}

By substitution, you will demonstrate that under perfect Airy compensation, the gravity effects of topography and root cancel exactly, and discuss what assumptions are required for this result.

\timeslot{Numerical Demonstration}{25}

The instructor presents a live Python demonstration illustrating:

\begin{itemize}
    \item Airy compensation with varying compensation fractions
    \item Pratt-style lateral density variations
    \item Resulting gravity anomalies and residuals
\end{itemize}

Interpretation of results is emphasised rather than coding details.

\timeslot{Group Synthesis}{25}

Each group prepares a brief summary addressing:

\begin{itemize}
    \item One reason crust-only Airy isostasy fails.
    \item One way lithospheric mantle structure resolves this failure.
    \item One equation linking topography, density, and gravity.
    \item One observation that could reduce non-uniqueness in isostatic interpretation.
\end{itemize}

\timeslot{Exit Questions}{10}

Individually, respond to the following:

\begin{enumerate}
    \item Identify one unrealistic assumption in either Airy or Pratt isostasy and explain why it fails.
    \item Explain how geological or geophysical constraints (e.g.\ seismology or petrology) help resolve ambiguity in isostatic models.
\end{enumerate}

\notepage
\notepage

\subimport{activities/}{activity_ocean_continent}
\subimport{activities/}{activity_isostatic_correction}