\begin{activity}{Numerical Demonstration: Isostatic Gravity Along a Cross Section}

\section*{Purpose}

In this demonstration, the instructor will use \texttt{isostasy\_gui.py} to
build a lithospheric cross section along a chosen topographic profile,
compute its gravity response under varying degrees of compensation and
layered lithospheric structure, and compare the predicted gravity to
observations. The purpose is to see the single-column reasoning from
Activities 5.A and 5.B applied continuously along a real profile, and to
meet the one genuinely new idea of the session: that compensation is not
all-or-nothing.

\section*{Learning Outcomes}

By the end of this demonstration, you should be able to:
\begin{itemize}
    \item Explain how the column-by-column mass balance from Activity 5.A extends to a continuous topographic profile.
    \item Describe what the degree of compensation $C$ represents, and predict how changing it should affect the amplitude of the predicted gravity anomaly.
    \item Explain how the choice of layered lithospheric structure (crust only vs.\ crust + mantle lithosphere) affects the gravity prediction for the same topography.
    \item Use misfit between predicted and observed gravity to reason about how well (and where) a profile is locally compensated.
    \item Identify what a computational model needs beyond what was required for the single-column calculations in 5.A and 5.B.
\end{itemize}

\subsection*{Background}

\paragraph{From columns to a profile.} Everything in this demonstration rests
on the same balance you used in Activity 5.A: at the depth of compensation,
column-integrated mass must match between neighboring columns. The only
difference here is scale --- instead of comparing two columns by hand, the
program applies this balance (or a partial version of it, below) at every
point along a chosen cross section, built from a loaded crustal thickness
model rather than two fixed values.

\paragraph{Degree of compensation.} So far, every problem has assumed
\emph{full} local compensation: the root exactly satisfies the mass balance
from 5.A. In reality, compensation is often partial, since lithospheric
strength allows some of a load to be supported elastically rather than
buoyantly (Activity 5.B, Extension). We describe this with a single
parameter $C$, where
\[
    r(x) = C\, r_{\mathrm{Airy}}(x), \qquad 0 \le C \le 1,
\]
and $r_{\mathrm{Airy}}(x)$ is the fully compensated root thickness from
Activity 5.B. $C=1$ recovers full local compensation; $C=0$ means the load
is entirely supported elastically, with no compensating root at all.

\paragraph{Gravity of a real profile.} The infinite slab formula from 5.B
assumed topography with no lateral edges. A real profile does not, so the
program instead sums the gravity effect of many finite blocks along the
cross section --- the same discrete 2-D prism approach you saw in Week~3
(Activity 3.B, Q11), not a new formula.

\subsection*{App Overview}

\begin{itemize}
    \item Select two points to define a cross section across the loaded topography.
    \item Load a crustal thickness model for the same profile.
    \item Specify a layered lithospheric model (layer densities and thicknesses), following the same structure as Activity 5.A Part B.
    \item Set the degree of compensation $C$ and compute the predicted gravity anomaly along the profile.
    \item Compare the predicted gravity to observed gravity along the same profile, if loaded.
\end{itemize}

\section*{Tasks}

Follow along as the instructor builds a cross section and adjusts each
parameter in turn. Answer the questions below as you go --- they are written
to apply to whichever profile is selected.

\begin{questions}

    \question{Full compensation baseline}
    With $C=1$, does the predicted gravity anomaly broadly track or oppose
    the topography along the profile? Explain why, referring back to what
    you showed in Activity 5.B, Part 3.
    \answerbox[3cm]

    \question{Varying the degree of compensation}
    As the instructor decreases $C$ from 1 toward 0, predict first, then
    observe: does the amplitude of the predicted gravity anomaly increase or
    decrease? Explain in terms of how much compensating root mass remains at
    each value of $C$.
    \answerbox[3cm]

    \question{Finding the best-fit compensation}
    If observed gravity is available for this profile, adjust $C$ to find
    the value that best matches the predicted and observed gravity. What
    does that value of $C$ tell you about how this section of lithosphere is
    actually being supported?
    \answerbox[3cm]

    \question{Model dependence}
    With $C$ held fixed, the instructor will change the assumed crustal
    density and/or reference lithospheric thickness. How does the predicted
    gravity change? Relate this to the "model-dependent correction" Key Idea
    from Activity 5.B.
    \answerbox[3cm]

    \question{Adding a layer}
    The instructor will now add or modify a mantle lithosphere layer, as in
    Activity 5.A Part B. Does the degree of compensation needed to fit the
    observed gravity change once this layer is included? Why might a
    two-layer model change your interpretation compared to a crust-only
    model?
    \answerbox[3cm]

    \question{Uniform compensation?}
    Does a single value of $C$ fit the observed gravity equally well across
    the whole profile, or does the fit degrade in some places (e.g.\ near a
    sharp topographic feature or a continent--ocean boundary)? What would
    that tell you about assuming one compensation state for an entire
    region?
    \answerbox[3cm]

    \question{What changed, computationally}
    Compared to the single-column, by-hand calculations in 5.A and 5.B, what
    additional ingredients did this model need to predict gravity along a
    real profile?
    \answerbox[3cm]

\end{questions}

\section*{Key Ideas}

\textbf{Compensation is a spectrum, not a switch.} Local isostatic balance
(as in 5.A and 5.B) and fully elastic support are two ends of a continuum;
most real topography is compensated to some intermediate degree, and that
degree can be estimated by comparing predicted and observed gravity.

\textbf{A single column-balance principle scales up without changing.} The
same mass balance used for two columns by hand applies, unchanged, at every
point along a continuous profile --- modelling a whole cross section is a
matter of scale and bookkeeping, not new physics.

\textbf{Fitting a model is itself an inference about the Earth.} The value
of $C$ (and the layered structure) that best matches observed gravity is
not just a curve-fitting exercise --- it is an estimate of how a real piece
of lithosphere is mechanically supported, with the same model-dependence
caveats that applied to the isostatic correction in 5.B.

\end{activity}